OpenMPIR

### 2017-09-07 paper.tex
Revise intro / continue to flesh out tapir

@@

When writing task-parallel programs today, one has a large selection of
programming models. Unfortunately, despite having significant overlap in
semantics, parallel programming models like OpenMP, Cilk, Kokkos, HPX, Charm++,
Qthreads, pthreads, MPI, Chapel, UPC, etc. have virtually no ability to
interoperate, both at compile time and run time. In addition, many of these
programming models are implemented as libraries, and therefore have no compiler
support. There has been some recent work on specializing compilers to reason
about parallel programming libraries \cite{Moss_2016}, but fragmentation among
models presents a significant challenge for compiler writers, which must choose
a single model to analyze and optimize. An ideal solution would be some form of
common representation of parallelism.

Compilers use internal representations \textbf{(IRs)} for implementing the
majority of their analyses and optimizations \cite{llvm}. Historically, a major
drawback for parallel programs is that IRs haven't had the ability to
represent, and therefore reason about, parallelism.  This has prevented the
application of common optimizations to concurrent code.  The recent work of
Schardl et al. \cite{tapir} has shown that real world compiler IRs can be
extended to represent fork-join parallelism. By extending the LLVM instruction
set with three instructions, they were able to capture all of the semantics to
implement the semantics of Cilk, an extension to C. They suggested that a
similar approach could be taken with OpenMP tasks.

In this work, we put that suggestion to the test, implementing OpenMP tasks by
compiling them to Tapir instructions. This enables us to target non-openmp
runtimes and compare performance to existing OpenMP implementations. For this
paper, our prototype implementation uses a Cilk runtime as a backend, and we run
our prototype implementation on the Barcelona OpenMP task suite
\cite{barcelona}. We discuss the kinds of optimizations this enables, and how
the work could be extended to include other parallelism constructs and semantics
in OpenMP. We also discuss the possibility of extending this work to other
programming models, which would help to alleviate some of the fragmentation
issues discussed above.

@@

\wmnote{This feels like we're trying to convince someone about LLVM: perhaps
LLVM has gained traction in the academic community for both its simplicity and
relatively painless modification} 
\gsnote{I wanted to give a brief nod to the importance of LLVM, given the venue
(LLVM in HPC). Is this any more satisfactory?} 
Among IRs, LLVM is arguably the most ubiquitous \cite{lattner2004llvm}. Winner
of the 2012 ACM Software System Award, it is in use by frontends ranging from
Haskell to C++, and backends ranging from x86 to PTX \cite{lattner2004llvm}.
Implementing a simple, extensible single static assignment IR, LLVM enables
powerful optimizations that are effective for a wide range of programming
languages and hardware \cite{lattner2004llvm}.

Historically, LLVM has had a sequential semantics: there is no explicit notion
of concurrent or parallel processes. This has resulted in existing compilers
treating parallelism constructs as thin wrappers for calls into runtime
systems. These calls are infeasible for the compiler to reason about, due to
the complexity and flexibility of runtimes being called. This situation was
recently remedied by Schardl et al.  \cite{tapir}. By adding three instructions
to LLVM, Tapir enables analyses and optimizations of parallel code. We describe
Tapir in detail in Section~\ref{Sec:Tapir}. For a more complete description see
\cite{tapir}.

@@

\setminted{fontsize=\small,baselinestretch=1}
\begin{figure}[t]
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l@{}l}

@@

\subsection{Tapir Internals}
Tapir is a compelling solution to this problem because it allows existing
analysis and optimizations to work on parallel programs without much modification.
Tapir does this by extending LLVM's instruction set to natively represent
fork-join parallel programs in a way that

... TODO ...

Specifically, Tapir introduces the \texttt{detach}, \texttt{reattach}, and
\texttt{sync} instructions. \wmnote{How in detail should I go here? I can talk
about these instructions on a broad level as a target for parallel tasks, or
more concretely define what they do. However, we don't seem discuss CFG's
before this so I'm not sure what level of introduction you're thinking of.}
\gsnote{I was thinking roughly a page should be enough. Maybe a summary of what
Tapir is followed by an example of an optimization it enables. I like both the
CSE and tail-call optimization examples from your PPoPP paper. Feel free to 
define CFG and any other terms you need to. If that's too much to ask on such 
short notice just let me know and I'll help out.}

@@

Recursive task creations and synchronizations using the constructs result in 
an implicit directed acyclic graph (DAG) that allows both reasoning about and 
visualization of the program execution.  Figure~\ref{Fig:fib-graph-to-schedule}
shows some example code, a view of the task DAG, and a simplified execution 
schedule mapping the tasks to a team of two threads.  In the example, the Nth 
Fibonacci number is calculated by recursively generating tasks to calculate 
the (N-1)th and (N-2)th Fibonacci numbers.  The \texttt{taskwait} ensures that 
the child tasks have completed before their answers are combined to yield the 
final result.

\begin{figure*}
\begin{center}
\begin{tabular}{| c c c | c c c | c c c |}
\hline
 & 
\begin{lstlisting}
int fib(int n) 
{                    // A
  if (n < 2)
    return n;
  else
  {
    int x, y;
    #pragma omp task
      x = fib(n-1);  // B
    #pragma omp task
      y = fib(n-2);  // C
    #pragma omp taskwait
    return (x+y);    // D
  }
}
\end{lstlisting}
& & &
    \centering{
   \raisebox{-0.6in}{\includegraphics[scale=0.35]{fib-task-graph.pdf}}
    }
& & &
    \centering{
     \raisebox{-0.25in}{\includegraphics[scale=0.35]{fib-schedule.pdf}}
    }
 &  \\
\hline
\end{tabular}
\end{center}
\caption{Code, task graph, and schedule of a simple brute force recursive 
Fibonacci number calculation on two threads.}
\label{Fig:fib-graph-to-schedule}
\end{figure*}

@@

To better understand how the OpenMP to Tapir compiler works, we turn to the
\texttt{fib} example from Section~\ref{Sec:OpenMP}. In Figure~\ref{Fig:Example}
we show what Tapir code is generated.

@@

\caption{Compilation to Tapir}
\label{Fig:Example}
\end{figure*}

@@

\section*{Acknowledgments}
Sandia National Laboratories is a multimission laboratory managed and operated 
by National Technology and Engineering Solutions of Sandia, LLC., a wholly 
owned subsidiary of Honeywell International, Inc., for the U.S. Department of 
Energy’s National Nuclear Security Administration under contract DE-NA-0003525.

@@

Work supported by the Advanced Simulation and Computing program of the
U.S. Department of Energy's NNSA\@.  Los Alamos National Laboratory is
managed and operated by Los Alamos National Security, LLC (LANS), under
contract number DE-AC52-06NA25396 for the Department of Energy’s
National Nuclear Security Administration (NNSA).

>>>>

\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing,shapes,arrows,positioning,calc}

\tikzset{
  basic block/.style={
    rectangle,
    rounded corners,
    draw=black, thin,
    % minimum width=3.2cm,
    inner sep=3pt,
    % outer sep=4pt,
    align=left,
    anchor=west,
    fill=codebgcolor, 
    % font={\ttfamily\fontsize{8.5}{10.2}\selectfont},
    font={\ttfamily\fontsize{7.5}{9}\selectfont},
    text=terminalcolor,
    draw=codeheadcolor,
    thick,
    auto,
    node distance=.4cm and 1.25cm,
  },
  cfedge/.style={
    font=\itshape,
    draw=black,
    ->,
    >=stealth'
  },
  process/.style={
    draw,
    fill=orange!50,
    rectangle,
    minimum height=1.5em,
    minimum width=5em,
    align=center,
    font=\small,
  },
  form/.style={
    draw,
    fill=blue!30,
    ellipse,
    minimum height=1em,
    align=center,
    font=\footnotesize,
  },
  slave/.style={
    execute at end picture={
      \coordinate (lower right) at (current bounding box.south east);
      \coordinate (upper left) at (current bounding box.north west);
      \pgfresetboundingbox
      \path (upper left) rectangle (lower right);
    }
  }
}

@@

When writing task-parallel programs today, one has a large selection of
programming models. Unfortunately, despite having significant overlap in
semantics, parallel programming models such as OpenMP\cite{}, Cilk\cite{},
Kokkos\cite{}, HPX\cite{}, Charm++\cite{},
Qthreads\cite{}, pthreads\cite{}, MPI\cite{}, Chapel\cite{}, UPC\cite{},
among others have virtually no ability to interoperate either at compile
time or run time.
\wmnote{Not sure if relevant, but there was some experiment run a while back of someone composing a library that internally used OpenMP
with another framework that had terribly results compared with either used alone.}
Many of these programming models are implemented as runtime libraries with only
primitive compiler support. As a result, most of these parallel frameworks
aren't able to take advantage of the compiler's analysis and optimizations.
There has been some recent work on specializing compilers to reason
about parallel programming libraries \cite{Moss_2016}, but fragmentation among
models presents a significant challenge for compiler writers, which must choose
a single model to analyze and optimize. An ideal solution would be some form of
common representation of parallelism.

Compilers implement the majority of their analyses and optimizations on
intermediate representations \textbf{(IRs)} which allow such transformations
to be written once with relative ease and apply to a variety of source languages
such as C or C++.
Historically, the IR of mainstream compilers such as LLVM\cite{llvm} or GCC's Gimple\cite{}
haven't suppored parallel programs. As a consequence, compilers relying on these
IRs haven't had the ability to
represent, and therefore reason about, parallelism. The recent work of
Schardl et al. \cite{tapir} has shown that real world compiler IRs can be
extended to represent fork-join parallelism without requiring a major rewrite.
They do so by extending the LLVM instruction set with three instructions
capable of representing fork-join parallelism. To benchmark performance and test
accuracy, they wrote a frontend which translated programs written with Cilk, a parallel
programming model for C/C++, into Tapir. They also wrote a backend to \textbf{lower}
take Tapir programs to vanilla LLVM with Cilk runtime calls.
They suggested that a similar approach could be taken with OpenMP and other
such frameworks.

In this work, we put that suggestion to the test, implementing OpenMP tasks by
compiling them to Tapir instructions. This allows us to both analyze and optimize
OpenMP programs as well as compile programs written in OpenMP to use other runtimes.
For this paper, our prototype implementation uses a Cilk runtime as a backend,
and we run our prototype implementation on the Barcelona OpenMP task suite
\cite{barcelona}. We discuss the kinds of optimizations this enables, and how
the work could be extended to include other parallelism constructs and semantics
in OpenMP. We also discuss the possibility of extending this work to other
programming models, which would help to alleviate some of the fragmentation
issues discussed above.

@@

\gsnote{I wanted to give a brief nod to the importance of LLVM, given the venue
(LLVM in HPC). Is this any more satisfactory?}
\wmnote{It still feels like we're trying to advocate for LLVM. I'd prefer some
more objective-sounding language like perhaps the one below.}
LLVM has gained traction in the academic community for its simplicity,
relatively painless modification, and extensibility\cite{lattner2004llvm}.
Historically, LLVM has had a sequential semantics: there is no explicit notion
of concurrent or parallel processes. This has resulted in existing compilers
treating parallelism constructs as thin wrappers for calls into runtime
systems. These calls are infeasible for the compiler to reason about, due to
the complexity and flexibility of runtimes being called. This situation was
recently remedied by Schardl et al. \cite{tapir}. By adding three instructions
to LLVM, Tapir enables analyses and optimizations of parallel code. We describe
Tapir in detail in Section~\ref{Sec:Tapir}. For a more complete description see
\cite{tapir}.

@@

\begin{figure}[t]
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l@{}l}

@@

\subsection{Tapir Internals}
Tapir is a compelling solution to this problem because it allows existing
analysis and optimizations to work on parallel programs without much modification.
To properly describe Tapir, one must briefly introduce key aspects of LLVM.
LLVM represents control flow through a control-flow graph (CFG) of blocks,
which are sequences of instructions. Blocks always end with terminator instructions
which define the edges of the control flow graph. For example, \figref{CFG} shows a simple
function and its corresponding LLVM code.

\begin{figure*}[h!]
  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}ll}

    \subfiglabel{a} & \subfiglabel{b} \\
\begin{minipage}[T]{0.45\linewidth}
      \begin{lstlisting}
int f(int x) {
  int y = 10;
  if (x == 0) {
    y = 12;
  }
  return y;
}
      \end{lstlisting}
    \end{minipage}
&
    \begin{minipage}[T]{0.45\linewidth}
\begin{lstlisting}[language=llvm]
define i32 @f(i32 %x) {
entry:
  %y = alloca i32, align 4
  store i32 10, i32* %y, align 4
  %cmp = icmp eq i32 %x, 0
  br i1 %cmp, label %if.then, label %if.end

if.then:
  store i32 12, i32* %y, align 4
  br label %if.end

if.end:
  %1 = load i32, i32* %y, align 4
  ret i32 %1
}
\end{lstlisting}
    \end{minipage}\\
    \addlinespace[2ex]
    \bottomrule
  \end{tabular*}
  \caption{Example code snippet and the corresponding LLVM code. This code segment contains three blocks: \code{entry}, \code{if.then}, and \code{if.end}. The conditional statement is implemented in LLVM by the conditional branch at the end of the entry block. }
  \figlabel{CFG}
  \vspace{-.4cm}
\end{figure*}

To represent parallelism, Tapir introduces the \code{detach}, \code{reattach}, and
\code{sync} instructions. The \code{detach} statement is syntactically similar to
the conditional branch as it is also a terminator instruction with with two blocks.
However, unlike the conditional branch which denotes a choice between successors, the \code{detach} instruction denotes that its two successors may run in parallel. Semantically,
the first successor of a \code{detach} instruction is referred to as the \textbf{detached block} and represents a task that can, but is not required to, run in parallel with the
\textbf{continuation block}, or the second successor to the detach statement. The \code{detach}
is analagous to a \code{fork} in most fork-join frameworks.

At the end of any task created by a \code{detach} instruction, one must end with a \code{reattach} instruction to the corresponding continuation block. The \code{reattach} instruction
is used to denote the end of a parallel task.

Finally, one may use the \code{sync} instruction to wait for any outstanding tasks created
by \code{detach} instructions within the current function to complete. The \code{sync}
instruction is analagous to a \code{join} in most fork-join frameworks.

@@

Recursive task creations and synchronizations using the constructs result in 
an implicit directed acyclic graph (DAG) that allows both reasoning about and 
visualization of the program execution.  \figref{fib-graph-to-schedule}
shows some example code, a view of the task DAG, and a simplified execution 
schedule mapping the tasks to a team of two threads.  In the example, the Nth 
Fibonacci number is calculated by recursively generating tasks to calculate 
the (N-1)th and (N-2)th Fibonacci numbers.  The \texttt{taskwait} ensures that 
the child tasks have completed before their answers are combined to yield the 
final result.

\begin{figure*}
\begin{center}
\begin{tabular}{| c c c | c c c | c c c |}
\hline
 & 
\begin{lstlisting}
int fib(int n) 
{                    // A
  if (n < 2)
    return n;
  else
  {
    int x, y;
    #pragma omp task
      x = fib(n-1);  // B
    #pragma omp task
      y = fib(n-2);  // C
    #pragma omp taskwait
    return (x+y);    // D
  }
}
\end{lstlisting}
& & &
    \centering{
   \raisebox{-0.6in}{\includegraphics[scale=0.35]{fib-task-graph.pdf}}
    }
& & &
    \centering{
     \raisebox{-0.25in}{\includegraphics[scale=0.35]{fib-schedule.pdf}}
    }
 &  \\
\hline
\end{tabular}
\end{center}
\caption{Code, task graph, and schedule of a simple brute force recursive 
Fibonacci number calculation on two threads.}
\label{fig:fib-graph-to-schedule}
\end{figure*}

@@

To better understand how the OpenMP to Tapir compiler works, we turn to the
\texttt{fib} example from Section~\ref{Sec:OpenMP}. In \figref{fig:Example}
we show what Tapir code is generated.

@@

\caption{Compilation to Tapir}
\label{fig:Example}
\end{figure*}

@@

\section*{Acknowledgments}
Sandia National Laboratories is a multimission laboratory managed and operated 
by National Technology and Engineering Solutions of Sandia, LLC., a wholly 
owned subsidiary of Honeywell International, Inc., for the U.S. Department of 
Energy's National Nuclear Security Administration under contract DE-NA-0003525.

@@

Work supported by the Advanced Simulation and Computing program of the
U.S. Department of Energy's NNSA\@.  Los Alamos National Laboratory is
managed and operated by Los Alamos National Security, LLC (LANS), under
contract number DE-AC52-06NA25396 for the Department of Energy's
National Nuclear Security Administration (NNSA).
